---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-", # Verificar si esto es lo que da problemas para la inserción de imágenes.
  out.width = "100%"
)
```

# ILS: An R package for statistical analysis in Interlaboratory Studies


<!-- badges: start -->
<!-- badges: end -->

## Overview

The main objective of the ILS package is to detect laboratories that provide not consistent
results, working simultaneously with different test materials, from the perspective
of the Univariate Data Analysis and the Functional Data Analysis (FDA).

The ILS package can identify laboratories that provide significantly different
results through the estimation the Mandel’s h and k scalar statistics, based
on the ASTM E691 and ISO 5725-2 standards. Additionally, the package has implemented
tools to assess the presence of outliers using the Cochran and Grubbs tests.

Furthermore, Analysis of Variance (ANOVA) techniques are provided, both for the
cases of fixed and random effects, including confidence intervals for the
principles parameters.

Moreover, one of the novelties of this package is the incorporation of
tools to perform an ILS from a functional data analysis approach. Accordingly, the
functional nature of the data obtained by experimental techniques corresponding
to analytical chemistry, applied physics and engineering applications (spectra,
thermograms, and sensor signals, among others) is taking into account by implementing
the functional extensions of Mandel’s h and k statistics.

<br>


## Concepts across the board

**ILS package** addresses several concepts and approaches which will be discussed
in detail in the later sections of this article. The most relevant concepts can be
found below:

-   **Interlaboratory Study (ILS)**

    This concept can be defined as a control procedure to
    evaluate the performance of a group of laboratories through a collaborative
    trials. In an Interlaboratory Study, an adequate number of laboratories are
    chosen to participate in the experiment with the aim of analyzing the samples
    and obtain results.
    
    Participating laboratories receive samples (previously homogenized or to be
    homogenized by the laboratories) for analysis, then, the measurements results of
    the laboratories are evaluated according to the degree of data variability. Some
    of the most common factors that may be a cause of variability are: the equipment
    of laboratories, operators, materials, temperature and humidity, among others.
    
-   **Repeatability and Reproducibility studies (R&R)**

    Repeatability and reproducibility (R&R) studies are a relevant methodology in
    quality control that assess the accuracy of a measurement system. In addition,
    R&R studies allow the identification of errors and sources of variation in the
    measurement process, and are essential to ensure the reliability and accuracy of the results.
    
    This tool of quality control focuses on two key aspects of measurement:
    
    **Repeatability:** Repeatability is the variation between successive measurements
    of the same part or trait by the same person using the same gage. In other words,
    how much variation do we see in measurements taken by the same person, on the same part, using the same tool.
    
    **Reproducibility:** Reproducibility is the difference in the average of the measurements
    made by different people using the same instrument when measuring the identical
    characteristics on the same part. In other words, how much variation do we see
    in measurements taken by different people on the same part using the same too.

-   **Consistency tests**

    Several scalar statistical techniques are frequently applied to study the consistency
    of test results from the different laboratories that participate in an ILS.
    Standard ASTM E-691 (Standard Practice for Conducting an Interlaboratory
    Study to Determine the Precision of a Test Method) recommends applying only
    one graphical technique from Mandel’s k and h statistics [2], while ISO 5725-2
    (Accuracy – trueness and precision – of measurement methods and results) recommends,
    in addition to the graphic technique, to use the Cochran and Grubbs.
    
    The 2 standards mentioned can be consulted in the following accesses:
    
    [ASTM E-691](https://www.astm.org/e0691-21.html)
    
    [ISO 5725-2](https://es.scribd.com/document/745647300/NTP-ISO-5725-2-2021-Antecedente-2019)

-   **ANOVA tests**

    The technique of  Analysis of Variance (ANOVA), the effect of the laboratory
    factor over the response can be studied. The variance of repeatability and
    reproducibility can be also estimated when an ANOVA random effects model
    is considered, as shown in ISO 5725-2 [1]. On the other hand, if a fixed effect
    model is fitted, in addition to the F test, multiple comparisons of means can be
    performed with the Tukey Honest Significant Difference (HSD) method.
    
    
-   **Functional Data Analysis (FDA)**:

    Functional Data Analysis is a relatively new branch of statistics that takes curves as unit of analysis,
    also surfaces, and volumes defined in a continuum (such as time, or frequency’s
    domain). Considering the recent advances in computing science, and the increasing
    amount of data generated by experimental techniques and sensors, the
    FDA has had a great development in recent years. In fact, we have many statistical
    methodologies that have been developed and extended to the functional
    case, such as exploratory analysis, regression, classification, analysis of variance,
    and time series.
    
    In the specific case of ILS, FDA extensions for Mandels’s
    h and k have been proposed and described by Flores, in addition to
    other works where the FDA descriptive analysis had been introduced for ILS
    studies.
    
<br>

## The R Package ILS

The present ILS library implements and calls some of their routines in order
to perform outlier detection in the framework of the Interlaboratory Studies.
Thus, regarding to ILS with scalar response, there are some interesting and
useful computational tools in R software.

Particularly, the metRology package estimates the uncertainty of the measurement, and
performs the required statistical calculations for Interlaboratory Studies, whereas
multcomp performs analysis of variance (ANOVA) through F and Tuckey tests.

On the other hand, due to the exponential increasing of FDA available techniques,
there are also a continuously increasing number of R libraries devoted to this branch
of statistics. Among all of them, the most important and used packages (on which the
present proposal is based) are fda.usc, that implements outlier detection
techniques and functional ANOVA, among other tools for FDA, and the fda. The present
ILS package uses the applications of the before mentioned multcomp and fda.usc

In summary, the functions incorporated to the ILS package to perform Interlaboratory Studies
are classified under the following scheme:

.......

<br>

## Interlaboratory Studies: Standard Approach

The ILS package provides two groups of functions made to detect outlying
individual results (outlying replicates) and outlying laboratories;  for this purpose, 
the package offers graphical and analytical procedures (statistical hypothesis test).

As mentioned in the previous section, among the methodologies used to evaluate the
consistency of laboratory results, we must highlight the r&R studies, which quantify
the variability between laboratories (reproducibility) and variability between results (repeatability).

-   **Mandel’s h and k statistic**

    Accordingly with the repeatability and reproducibility concepts, this statistics are used
    in ILS to detect laboratories that provide inconsistent results. The h statistic
    explains the variability between the laboratories, that is,
    estimates the bias, which is the difference of the means of each laboratory with
    respect to the global mean, while the k statistic estimates the variability within
    the laboratories, comparing the repeatability corresponding to each laboratory.
    
    The decision rule for detecting whether a laboratory is inconsistent is based
    on the comparison between the value of the h or k statistic and the critical value,
    which is normally calculated with a significance level of 0.5% (as recommended
    by ASTM E-691).

-   **Cochran and Grubbs test**

    The aim oh these tests is to examine the consistency within a laboratory, whereas
    the Grubbs test is commonly used to examine consistency between laboratories.
    The Grubbs test can also be used as a consistency test for the results obtained
    in a laboratory using identical materials. These tests are recommended by ISO 5725-2.
    
Accordingly, consistency tests and identification of atypical results must be performed previously to ANOVA
analysis. There are two possible scenarios in which the presence of outliers
can be evaluated: the first is that the results of one laboratory deviates from
the others in terms of precision, that is, when the measurements made by a
laboratory significantly differ with respect to the measurements obtained by
other laboratories. The second scenario is related with the identification of
outliers in a laboratory for a certain level.

<br>

The statistics and tests recommended by ISO 5725-2 and ASTM E-691 are described below.

-   **Mandel’s h statistic and Grubbs test**

    If a laboratory is identified as an outlier, after applying the h statistic and
    the Grubbs test to different levels within a laboratory, this is an evidence of the
    presence of a laboratory high bias (due to a high systematic error in calibration,
    or errors in the equations used to compute the results).

-   **Mandel’s k statistic and Cochran test**

    The Cochran test only evaluates the highest value in a series of variances. If
    a laboratory is detected as an outlier, using the k statistic or with the Cochran
    test, this indicates that the variance within the laboratory is high (due to lack
    of familiarity with the analytical method, differences of appreciation among operators,
    inadequate equipment, equipment in poor state, or careless execution),
    in which case the total of results collected by this laboratory should be rejected
    and taken out of the study.
    
    The detection of inconsistent laboratories must be repeated until it stops
    reporting outliers. However, the consistency tests should be used with caution,
    because if this process is carried out in excess, could lead to false outlier
    identification.

<br>

## Interlaboratory Studies: New FDA Approach

The ILS package have some techniques developed to check if inconsistent laboratories
are detected either by outliers in the within-laboratory or in between-laboratory variability.

-   **Hypothesis of reproducibility and repeatability**:

     To test the hypothesis that there are no differences in the measurements between
     the laboratories in a study, functional statistics Hl(t) and Kl(t) are estimated
     for each laboratory.
     
     The null hypothesis of reproducibility states that
     
     .......
     
     To evaluate the reproducibility of the laboratory results, the H(t) statistic is
     calculated as follows:
     
     .......
     
     
     To test
     the reproducibility hypothesis, the test statistic dH
     
     .......
     
     On the other hand, the null hypothesis of repeatability is:
     
     .......
     
     The repeatability test is based on the K(t) statistic, expressed as:
     
     .......
     
     The test statistic is dK.
     
<br>

A bootstrap procedure to reproduce the distribution of these statistics under
the corresponding null hypothesis is proposed in ... This procedure provides
the approximation of the critical values cK and cH of the test statistics dH and dK,
respectively. Additionally, confidence bands for the functional statistics
H(t) and K(t) can be computed (determined by the envelope of their bootstrap
counterparts with a norm less than the corresponding critical value).

<br>


## ILS package Installation

You can install the development version of ILSR6V1 from [GitHub](https://github.com/) with:

``` r
# install.packages("ILSR6V1")
library("ILSR6V1")
```

<br>


## ILS package Practical examples


## 1) Clinical study of blood glucose measurement

<br>

-   **Graphical and analytical statistics**

The Glucose dataset corresponds to the serum glucose test (measurements
of the concentration of glucose in the blood used to control the diabetes). In this
study, eight laboratories where involved, and five different tests were performed
on blood samples labelled with different references, ranging from a low sugar
content to a very high one. Three replicates were obtained for each sample.

Each of these laboratories measured five different concentration levels (A, B,
C, D, E) of a given material, and at each of these levels, three measurements
were taken (3 replicates). Each laboratory provided a total of 15 measurements
(3 for each level), therefore, with 8 laboratories involved, 120 measurements
were obtained.

In order to access this dataset, the ILS package installing and loading is
required. Once loading is performed, the Glucose data.frame object is called
using the following instructions.

```{r example_1}
suppressWarnings(suppressMessages(library(ILS)))
data("Glucose")
```
<br>

The first step to perform an analysis with the ILS package consist on using
the function ils.qcdata() (quality control data) that receives a data.frame
as an argument. By default, the first column of the data frame must contain
the response variable, the second column the replicates, the third column corresponds
to the tested material, while the fourth column includes the laboratories
where the procedure was performed.

For instance, the following code creates an object qcdata (of class ils.qcdata)
and uses the summary() method to obtain descriptive statistics information of
the dataset.

```{r example_2}
qcdata <- lab.qcdata(Glucose)
summary(qcdata)

plot(qcdata,
     ylab = "Laboratory",
     xlab = "Glucose concentration in blood")
```

The figure shows the obtained values for each laboratory and material. It can
be noted that the blood glucose level increases from material A to D and there
is more variability between the results for each laboratory from material C to
material E.

<br>

In order to calculate the graphical and analytical statistics for the scalar
(univariate) case, first, the function lab.qcs() (quality control statistics) has to
be used. This function returns the estimates of the statistical required measures
(mean, variance, etc.) for computing the Mandel’s h and k statistics, as well as
the required measures to perform the Cochran and Grubbs tests.

In the following commands, the lab.qcs() function uses the qcdata object
to create the qcstat object that estimates both the mean and the global deviation
 from the results of all laboratories and all materials. The repeatability
deviation (Sr), the deviation between the means of laboratories (SB), and the
reproducibility deviation (SR) for each material are also estimated (following
ASTM E-691)

```{r example_3}
qcstat <- lab.qcs(qcdata)
summary(qcstat)

plot(qcstat,
     xlab = "Deviations",
     ylab = "Material")
```
In the figure presented, the values of S (the global deviation of all laboratories), Sr
(the repeatability’s deviation), SR (reproducibility’s deviation) and SB (the
deviation between the means of the laboratories) are shown for each material.
A greater variability can be noted from material C to material E. Materials C
and D have a greater variability between the results of the laboratories (SR)
and within them (Sr).

<br>

-   **Consistency tests**

First, an analysis of the variability for
each laboratory will be performed. For this purpose, the k statistic (k.qcs())
and the Cochran test (cochran.test()) will be used to identify if there is any
laboratory with non-consistent results. Subsequently, the h statistic (h.qcs())
and the Grubbs test (grubbs.test()) will be used to perform an analysis to
evaluate inter-laboratory variability.

The following code creates an object containing the scalar Mandel’s k statistic
for each laboratory and material, which is subsequently plotted:

```{r example_4}
k <- k.qcs(qcdata, alpha = 0.005)
plot(k)
```

<br>
<br>

In the figure above, the dotted line represents the critical value equal to 2.06,
obtained for p = 8, n = 15 and α = 0.005 (following the ASTM standard). Hence,
outliers were detected for the laboratory 2, when testing the material 5, and for
material 3, when is tested by the laboratory 4, since the corresponding values
of the k statistics were greater than the critical value.

The summary() method prints the violations matrix. If an entry of this matrix
is FALSE, the laboratory reports outlying results for the corresponding material
at the specified significance level.

```{r example_5}
summary(k)
```

The Cochran’s test can be performed with the cochran.test() function.
In this case study, with the maximum variance for each material, no laboratory
was considered inconsistent, since the critical value was 0.52 and the p-values
for each material exceeded the 5% significance level.

```{r example_6}
cochran.test(qcdata)
```

We proceeded to use the methods h.qcs() and plot() to estimate and plot
the h statistics for each laboratory and material. The critical value was 2.15,
therefore, from this result it can be seen in Figure 6 that laboratories 4, 7 and
8 presented non-consistent results at a significance level of α = 0.005.

```{r example_7}
h <- h.qcs(qcdata, alpha = 0.005)
plot(h)


summary(h)
```

<br>

-   **Outliers detection**

Laboratories with very extreme results were detected by using the
Grubbs’ test, i.e. laboratories defined by very large or small results (glucose
content).

```{r example_8}
grubbs.test(qcdata)
```

Once the outlier laboratories are removed, one-way ANOVA analysis and
mean comparison test can be performed. The ILS package provides function
ils.aov() to perform an analysis of variance considering random or fixed effects
of the laboratory factor depending on the value of the random argument. By
setting this parameter to TRUE, a random effects one-way ANOVA is done for
each material, providing the corresponding F tests and confidence intervals for
the variances. For instance, the following code shows an example using the
results of one material for the sake of simplicity.

```{r example_9}
Glucose2 <- subset(Glucose, Material == "A")
Glucose2 <- lab.qcdata(Glucose2)
lab.aov(Glucose2, random = TRUE, level = 0.95)
```

<br>

## 2) Characterization of materials by thermogravimetric analysis

<br>

-   **Graphical and functional analytical statistics**

This exmample is about 7 (emulated) laboratories that analysed 15 samples of calcium oxalate by
Thermogravimetric (TG) techniques, obtaining 105 TG curves that shows the
mass loss of oxalate as a function of temperature when the oxalate samples were
heated at 20◦C/min. The TG dataset contains the resulting values at 1000 discretization points,
corresponding to temperatures ranging from 40◦C to 850◦C.
Laboratories 1, 6, and 7 presented non-consistent results. In laboratory 1 a Simultaneous
Thermal Analyzer (STA) was used with an out of phase calibration
program. In laboratory 6, we used a simultaneous SDT analyser with an old
calibration, and finally, in laboratory 7, we used a simultaneous SDT analyser
with a bias in the the temperature calibration with respect to the real values
(2◦C displaced with respect to the melting point of the zinc).

From the TG curves, a second set of data called IDT was obtained. The IDT
(Initial Decomposition Temperature) is a parameter defined by the temperature
at which the studied material losses the 5% of its weight when it is heated at a
constant rate. The dataset is composed of the IDT values of the calcium oxalate
samples analyzed by the 7 laboratories. It is an example of ILS study with scalar
response, obtained by extracting just only one representative feature from the
TG curve. It is important to stress that when a feature extraction process is
performed, there is the risk of loosing relevant information and thus obtaining
erroneous findings.

For the estimation of the functional statistics (for the performance of the
graphical and analytical methods), the procedure is the same as for the scalar
case. The ILS package provides the ils.fqcdata() function to generate the
functional quality control data from a matrix (or a data.frame), in which each
row represents a test result. The dimension of the matrix must be n×m, where n
is the number of replicates performed by the laboratories that participate in the
study, and m is the number of points observed in each curve. The usual methods
were implemented to make plots and summaries of the resulting objects.

<br>

To create an object that contains the functional quality control data from
the TG dataset, first we defined the grid in which the observations were obtained.
In this case, the grid consists of 1000 temperature values ranging from 40◦C to
850◦C.

```{r example_10}
data(TG)
delta <- seq(from = 40, to = 850, length.out = 1000)
fqcdata <- ils.fqcdata(TG, p = 7, argvals = delta)
plot(x = fqcdata,
     main = "TG curves of calcium oxalate",
     xlab = "Temperature (C)",
     ylab = "Mass (%)",
     legend = TRUE, x.co = 20, y.co = 90)
```

In the figure above, the TG curves are presented. The following commands create
a fqcstat object containing the functional statistics from fqcdata, and plot it
afterwards.

<br>

```{r example_11}
fqcstat <- ils.fqcs(fqcdata)
plot(fqcstat, xlab = "Temperature (C)", ylab = "Mass (%)")
```

The result is shown in this Figure. The plot() method creates a panel with four
graphs, in the first row the functional means and variances for each laboratory
are presented, while in the second row the global functional mean and variance
are plotted.

<br>

-   **Consistency tests**

With the aim to perform a r&R study of the TG and DSC functional datasets described
in the previous section, the corresponding statistics H(t), K(t), dH and dK will be estimated
and graphically represented. As above mentioned, the TG dataset contains thermogravimetric
test results from 7 laboratories, while the DSC dataset contains results from 6 laboratories
(excluding laboratory 1). Laboratories 1, 5 and 6 have provided different results from the
remaining laboratories and should be detected as outliers.

Function mandel.fqcs() estimates the functional statistics Hl(t) and Kl(t),
and the test statistics dKl and dHl, l = 1 . . . p. It also performs the bootstrap
procedure described in [6] to approximate the cK and cH critical values, and to
compute functional confidence bands for H(t) and K(t), under the corresponding null hypothesis. 
The results can be plotted with the default method.

```{r example_12}
set.seed(777)
mandel.tg <- mandel.fqcs(fqcdata, nb = 100)
plot(mandel.tg)
```

The left panels of the figure show the dKl and dHl test statistics for each
laboratory, l = 1 . . . p, as well as the corresponding critical values cH and cK
(horizontal lines), constructed at a significance level of α = 0.01. In the case
of the dH statistic, it is concluded that laboratory 7 does not meet the reproducibility
hypothesis (according with [6], laboratories 1, 6, and 7 were detected
as outliers through an iterative process). When the repeatability hypothesis is
tested by the dK test statistic, laboratory 6 is identified as an outlier in the first iteration.

The right panels of the figure show the functional statistics Hl(t) and Kl(t)
for each laboratory, l = 1 . . . p, as well as the corresponding confidence bands
(dotted lines in black). Functional estimates with a significant large test statistic
are shown with a solid line. These functional estimates allow us to identify in
which time intervals the laboratories provide different results. In this case,
the differences between the laboratory 7 and the remaining are in the interval
corresponding to the first degradation process of calcium oxalate.

<br>

-   **Outliers detection**

Finally, the ILS package is used to perform outlier detection tasks in the
Interlaboratory Study defined by the DSC dataset.

```{r example_13}
data(DSC)
fqcdata.dsc <- ils.fqcdata(DSC, p = 6,
                           index.laboratory = paste("Lab", 2:7),
                           argvals = delta)
mandel.dsc <- mandel.fqcs(fqcdata.dsc,
                          nb = 100)
plot(mandel.dsc)
```

The figure shows that repeatability hypothesis is not reject. Otherwise, the reproducibility
hypothesis is rejected, and laboratory 6 is properly detected as an outlier (see top panels
of the figure).

<br>

## Conclusions

The ILS package has been implemented in R software with the aim to provide the main
descriptive and outlier detection tools dealing with the Interlaboratory Studies and
recommended by the ISO 5725-4-1994 and ASTM E-691 standards. Namely, Mandels’s h
and k test (including their graphical output), Grubbs’ test, Cochran’s test, in addition
to ANOVA utilities.

Moreover, apart from standard scalar statistical techniques, the ILS
package provides FDA techniques to deal with functional data (curves), as data
unit, in the framework of Interlaboratory Studies. Indeed, the main novelty of
this computational proposal is the implementation of the functional extensions
of the h and k Mandel’s statistics when data results are curves, preventing to lose
relevant information derived from reduction dimension and feature extraction
processes. These new methods can identify successfully the outlier laboratories
directly from data curves.

